### 2022-06-13 doc/src/CSRs.tex

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name      & Description \\
\hline
\hline
\multicolumn{4}{|c|}{Machine-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x350} & MRW & \z{miselect} & Machine indirect register select \\
\z{0x351} & MRW & \z{mireg}    & Machine indirect register alias \\
\hline
\multicolumn{4}{|c|}{Machine-Level Interrupts} \\
\hline
\z{0xFB0} & MRO & \z{mtopi}    & Machine top interrupt \\
\hline
\multicolumn{4}{|c|}{Machine-Level IMSIC Interface (only with an IMSIC)} \\
\hline
\z{0x358} & MRW & \z{mseteipnum}
                   & Machine set external interrupt-pending bit by number \\
\z{0x359} & MRW & \z{mclreipnum}
                   & Machine clear external interrupt-pending bit by number \\
\z{0x35A} & MRW & \z{mseteienum}
                   & Machine set external interrupt-enable bit by number \\
\z{0x35B} & MRW & \z{mclreienum}
                   & Machine clear external interrupt-enable bit by number \\
\z{0x35C} & MRW & \z{mtopei}   & Machine top external interrupt \\
\hline
\multicolumn{4}{|c|}{Virtual Interrupts for Supervisor Level} \\
\hline
\z{0x308} & MRW & \z{mvien}    & Machine virtual interrupt enables \\
\z{0x309} & MRW & \z{mvip}     & Machine virtual interrupt-pending bits \\
\hline
\multicolumn{4}{|c|}{%
  Machine-Level High-Half CSRs for Interrupts 32--63 (RV32 only)} \\
\hline
\z{0x313} & MRW & \z{midelegh} & Extension of \z{mideleg} (only with S-mode) \\
\z{0x314} & MRW & \z{mieh}     & Extension of \z{mie} \\
\z{0x318} & MRW & \z{mvienh}   & Extension of \z{mvien} (only with S-mode) \\
\z{0x319} & MRW & \z{mviph}    & Extension of \z{mvip} (only with S-mode) \\
\z{0x354} & MRW & \z{miph}     & Extension of \z{mip} \\
\hline
\end{tabular}
\end{center}
\caption{Machine-level CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-M}
\end{table*}

@@

@@

Five CSRs, \z{mseteipnum}, \z{mclreipnum}, \z{mseteienum},
\z{mclreienum}, and \z{mtopei}, interact with an IMSIC, so exist only
when an IMSIC is implemented.
Like the indirectly accessed IMSIC registers, these CSRs are documented
in Chapter~\ref{ch:IMSIC}.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name      & Description \\
\hline
\hline
\multicolumn{4}{|c|}{%
  Supervisor-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x150} & SRW & \z{siselect} & Supervisor indirect register select \\
\z{0x151} & SRW & \z{sireg}    & Supervisor indirect register alias \\
\hline
\multicolumn{4}{|c|}{Supervisor-Level Interrupts} \\
\hline
\z{0xDB0} & SRO & \z{stopi}    & Supervisor top interrupt \\
\hline
\multicolumn{4}{|c|}{Supervisor-Level IMSIC Interface (only with an IMSIC)} \\
\hline
\z{0x158} & SRW & \z{sseteipnum}
                 & Supervisor set external interrupt-pending bit by number \\
\z{0x159} & SRW & \z{sclreipnum}
                 & Supervisor clear external interrupt-pending bit by number \\
\z{0x15A} & SRW & \z{sseteienum}
                 & Supervisor set external interrupt-enable bit by number \\
\z{0x15B} & SRW & \z{sclreienum}
                 & Supervisor clear external interrupt-enable bit by number \\
\z{0x15C} & SRW & \z{stopei}   & Supervisor top external interrupt \\
\hline
\multicolumn{4}{|c|}{%
  Supervisor-Level High-Half CSRs for Interrupts 32--63 (RV32 only)} \\
\hline
\z{0x114} & SRW & \z{sieh}     & Extension of \z{sie} \\
\z{0x154} & SRW & \z{siph}     & Extension of \z{sip} \\
\hline
\end{tabular}
\end{center}
\caption{Supervisor-level CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-S}
\end{table*}

@@

Register \z{stopi} reports the highest-priority interrupt that
is pending and enabled for supervisor level, as specified in
Section~\ref{sec:stopi}.
The IMSIC CSRs, \z{sseteipnum}, \z{sclreipnum}, \z{sseteienum},
\z{sclreienum}, and \z{stopei}, are described with the IMSIC in
Chapter~\ref{ch:IMSIC}.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name        & Description \\
\hline
\hline
\multicolumn{4}{|c|}{%
  Virtual Interrupts and Interrupt Priorities for VS Level} \\
\hline
\z{0x608} & HRW & \z{hvien}      & Hypervisor virtual interrupt enables \\
\z{0x609} & HRW & \z{hvictl}     & Hypervisor virtual interrupt control \\
\z{0x646} & HRW & \z{hviprio1}   & Hypervisor VS-level interrupt priorities \\
\z{0x647} & HRW & \z{hviprio2}   & Hypervisor VS-level interrupt priorities \\
\hline
\multicolumn{4}{|c|}{VS-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x250} & HRW & \z{vsiselect}
                   & Virtual supervisor indirect register select \\
\z{0x251} & HRW & \z{vsireg}
                   & Virtual supervisor indirect register alias \\
\hline
\multicolumn{4}{|c|}{VS-Level Interrupts} \\
\hline
\z{0xEB0} & HRO & \z{vstopi}     & Virtual supervisor top interrupt \\
\hline
\multicolumn{4}{|c|}{VS-Level IMSIC Interface (only with an IMSIC)} \\
\hline
\z{0x258} & HRW & \z{vsseteipnum}
                   & VS set external interrupt-pending bit by number \\
\z{0x259} & HRW & \z{vsclreipnum}
                   & VS clear external interrupt-pending bit by number \\
\z{0x25A} & HRW & \z{vsseteienum}
                   & VS set external interrupt-enable bit by number \\
\z{0x25B} & HRW & \z{vsclreienum}
                   & VS clear external interrupt-enable bit by number \\
\z{0x25C} & SRW & \z{vstopei}    & Virtual supervisor top external interrupt \\
\hline
\multicolumn{4}{|c|}{Hypervisor and VS-Level High-Half CSRs (RV32 only)} \\
\hline
\z{0x613} & HRW & \z{hidelegh}   & Extension of \z{hideleg} \\
\z{0x618} & HRW & \z{hvienh}     & Extension of \z{hvien} \\
\z{0x655} & HRW & \z{hviph}      & Extension of \z{hvip} \\
\z{0x656} & HRW & \z{hviprio1h}  & Extension of \z{hviprio1} \\
\z{0x657} & HRW & \z{hviprio2h}  & Extension of \z{hviprio2} \\
\z{0x214} & HRW & \z{vsieh}      & Extension of \z{vsie} \\
\z{0x254} & HRW & \z{vsiph}      & Extension of \z{vsip} \\
\hline
\end{tabular}
\end{center}
\caption{Hypervisor and VS CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-hypervisor}
\end{table*}

The VS CSRs in the table (\z{vsiselect}, \z{vsireg}, \z{vstopi},
\z{vsseteipnum}, \z{vsclreipnum}, \z{vsseteienum}, and
\z{vsclreienum}) all match supervisor CSRs, and substitute for those
supervisor CSRs when executing in a virtual machine (in \mbox{VS-mode}
or \mbox{VU-mode}).

@@

Along the same lines, when \z{hstatus}.VGEIN is not the number of
an implemented guest external interrupt, attempts from \mbox{M-mode}
or \mbox{HS-mode} to access CSRs \z{vsseteipnum}, \z{vsclreipnum},
\z{vsseteienum}, or \z{vsclreienum} raise an illegal instruction
exception, and attempts from \mbox{VS-mode} to access \z{sseteipnum},
\z{sclreipnum}, \z{sseteienum}, \z{sclreienum}, or \z{stopei} raise a
virtual instruction exception.

>>>>

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name      & Description \\
\hline
\hline
\multicolumn{4}{|c|}{Machine-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x350} & MRW & \z{miselect} & Machine indirect register select \\
\z{0x351} & MRW & \z{mireg}    & Machine indirect register alias \\
\hline
\multicolumn{4}{|c|}{Machine-Level Interrupts} \\
\hline
\z{0x35C} & MRW & \z{mtopei}   & Machine top external interrupt
                                  (only with an IMSIC) \\
\z{0xFB0} & MRO & \z{mtopi}    & Machine top interrupt \\
\hline
\multicolumn{4}{|c|}{Virtual Interrupts for Supervisor Level} \\
\hline
\z{0x308} & MRW & \z{mvien}    & Machine virtual interrupt enables \\
\z{0x309} & MRW & \z{mvip}     & Machine virtual interrupt-pending bits \\
\hline
\multicolumn{4}{|c|}{%
  Machine-Level High-Half CSRs for Interrupts 32--63 (RV32 only)} \\
\hline
\z{0x313} & MRW & \z{midelegh} & Extension of \z{mideleg} (only with S-mode) \\
\z{0x314} & MRW & \z{mieh}     & Extension of \z{mie} \\
\z{0x318} & MRW & \z{mvienh}   & Extension of \z{mvien} (only with S-mode) \\
\z{0x319} & MRW & \z{mviph}    & Extension of \z{mvip} (only with S-mode) \\
\z{0x354} & MRW & \z{miph}     & Extension of \z{mip} \\
\hline
\end{tabular}
\end{center}
\caption{Machine-level CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-M}
\end{table*}

@@

CSR \z{mtopei} also exists only when an IMSIC is implemented,
so is documented in Chapter~\ref{ch:IMSIC}
along with the indirectly accessed IMSIC registers.

@@

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name      & Description \\
\hline
\hline
\multicolumn{4}{|c|}{%
  Supervisor-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x150} & SRW & \z{siselect} & Supervisor indirect register select \\
\z{0x151} & SRW & \z{sireg}    & Supervisor indirect register alias \\
\hline
\multicolumn{4}{|c|}{Supervisor-Level Interrupts} \\
\hline
\z{0x15C} & SRW & \z{stopei}   & Supervisor top external interrupt
                                  (only with an IMSIC) \\
\z{0xDB0} & SRO & \z{stopi}    & Supervisor top interrupt \\
\hline
\multicolumn{4}{|c|}{%
  Supervisor-Level High-Half CSRs for Interrupts 32--63 (RV32 only)} \\
\hline
\z{0x114} & SRW & \z{sieh}     & Extension of \z{sie} \\
\z{0x154} & SRW & \z{siph}     & Extension of \z{sip} \\
\hline
\end{tabular}
\end{center}
\caption{Supervisor-level CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-S}
\end{table*}

@@

CSR \z{stopei} is described with the IMSIC in Chapter~\ref{ch:IMSIC}.

Register \z{stopi} reports the highest-priority interrupt that
is pending and enabled for supervisor level, as specified in
Section~\ref{sec:stopi}.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|l|l|l|l|}
\hline
Number & Privilege & Name        & Description \\
\hline
\hline
\multicolumn{4}{|c|}{%
  Virtual Interrupts and Interrupt Priorities for VS Level} \\
\hline
\z{0x608} & HRW & \z{hvien}      & Hypervisor virtual interrupt enables \\
\z{0x609} & HRW & \z{hvictl}     & Hypervisor virtual interrupt control \\
\z{0x646} & HRW & \z{hviprio1}   & Hypervisor VS-level interrupt priorities \\
\z{0x647} & HRW & \z{hviprio2}   & Hypervisor VS-level interrupt priorities \\
\hline
\multicolumn{4}{|c|}{VS-Level Window to Indirectly Accessed Registers} \\
\hline
\z{0x250} & HRW & \z{vsiselect}
                   & Virtual supervisor indirect register select \\
\z{0x251} & HRW & \z{vsireg}
                   & Virtual supervisor indirect register alias \\
\hline
\multicolumn{4}{|c|}{VS-Level Interrupts} \\
\hline
\z{0x25C} & HRW & \z{vstopei}    & Virtual supervisor top external interrupt
                                    (only with an IMSIC) \\
\z{0xEB0} & HRO & \z{vstopi}     & Virtual supervisor top interrupt \\
\hline
\multicolumn{4}{|c|}{Hypervisor and VS-Level High-Half CSRs (RV32 only)} \\
\hline
\z{0x613} & HRW & \z{hidelegh}   & Extension of \z{hideleg} \\
\z{0x618} & HRW & \z{hvienh}     & Extension of \z{hvien} \\
\z{0x655} & HRW & \z{hviph}      & Extension of \z{hvip} \\
\z{0x656} & HRW & \z{hviprio1h}  & Extension of \z{hviprio1} \\
\z{0x657} & HRW & \z{hviprio2h}  & Extension of \z{hviprio2} \\
\z{0x214} & HRW & \z{vsieh}      & Extension of \z{vsie} \\
\z{0x254} & HRW & \z{vsiph}      & Extension of \z{vsip} \\
\hline
\end{tabular}
\end{center}
\caption{Hypervisor and VS CSRs added by the Advanced Interrupt Architecture.}
\label{tab:CSRs-hypervisor}
\end{table*}

The VS CSRs in the table (\z{vsiselect}, \z{vsireg},
\z{vstopei}, \z{vstopi}, and, for RV32, \z{vsieh} and \z{vsiph})
all match supervisor CSRs, and substitute for those
supervisor CSRs when executing in a virtual machine (in \mbox{VS-mode}
or \mbox{VU-mode}).

@@

Along the same lines, when \z{hstatus}.VGEIN is not the number of
an implemented guest external interrupt, attempts from \mbox{M-mode}
or \mbox{HS-mode} to access CSR \z{vstopei}
raise an illegal instruction exception,
and attempts from \mbox{VS-mode} to access \z{stopei} raise a
virtual instruction exception.
